---
title: "WADDL to CSV"
author: "Brian High"
date: "2023-04-24"
output:
  html_document:
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 100), tidy = TRUE)
```

## Read WADDL file and convert to CSV

We will use an example "text" data file, a modified subset of a file we were told 
came from the [Washington Animal Disease Diagnostic Laboratory](https://waddl.vetmed.wsu.edu/ (WADDL). We were asked to read the file into R, but we were not told anything 
about the file format or structure, except that it was a "text" file.

This particular file format is challenging because the functions you would normally 
use for reading "text" files will not work without using some special options. 
Further, the data are not arranged in a familiar and [tidy](https://tidyr.tidyverse.org/articles/tidy-data.html) "columns are variables" structure.

We will explore how to address the issues that come up when reading this file: 

- Unexpected "embedded nulls" warning
- Unexpected "invalid multibyte string" error
- Unexpected file encoding type
- Unexpected delimiter (tab)
- Missing "elements" (values)
- Missing header
- Columns with no values (to remove)
- Rows missing any useful data (to remove)
- Unusual data structure (not "columns are variables")

For the curious, we also provide more details in three appendices:

- Appendix I: What's an "embedded null"?
- Appendix II: What's UTF-16LE?
- Appendix III: Good nulls and bad nulls

## Setup

First, we will load our R packages, set options, and set file paths.

```{r setup}
# Attach packages, installing as needed
if(!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(formatR, knitr, readr, dplyr, tidyr, stringr, purrr)

# Set options
options(readr.show_col_types = FALSE, readr.show_progress = FALSE)

# Set file paths
txt_file <- file.path("data", "waddl_data_example.txt")
csv_file <- file.path("data", "waddl_data_example.csv")
```

## Embedded Nulls

We expect a normal "text" file, because of the file suffix: `.txt`. If we knew it 
was a CSV file, we would use `read.csv()`, but from the `.txt` suffix we do not 
know the delimiter. It could be a comma, tab, space, or just about anything. So, 
we will start by trying `read.table()`. 

But when we read it in we get some unexpected messages.

```{r embedded_nuls_and_multibyte_strings}
# Try reading with read.table, just the first row
try(read.table(txt_file, nrows = 1))
```

The warning "appears to contain embedded nulls" will go away if we use `skipNul = TRUE`.

```{r skipnul}
# Try reading with read.table, just the first row
try(read.table(txt_file, nrows = 1, skipNul = TRUE))
```

## Invalid multibyte strings

The "invalid multibyte string" error is a clue that the file is not encoded as 
expected (usually UTF-8 or ASCII), so we check to see what it might be instead.

```{r encoding}
# Check file encoding
readr::guess_encoding(txt_file, n_max = 1)
```

And the guesses are listed with the most likely encoding at the the top of the 
list with a probability of 1. We can verify that by checking the first four bytes 
for the "byte order mark" (BOM):

```{r BOM}
# Look for a byte order mark (BOM) in the first four bytes to verify encoding
readBin(txt_file, what = "raw", n = 4)
```

We see a two-byte BOM of `ff fe`, followed by the first two-byte character of 
our data, "1" (the row number?), and "1" is `31 00` in UTF-16LE encoding. This 
verifies UTF-16LE encoding.

- See: https://www.unicode.org/faq/utf_bom.html#bom4
- And: https://en.wikipedia.org/wiki/Byte_order_mark#UTF-16

The BOM is the source of the "invalid multibyte string" error because `read.table()` 
does not know this is a UTF-16LE file having the associated BOM. 

The `\xff` and `\xfe` are the hexadecimal values for these two special "bytes", 
where a byte is 16 "bits". 

When we set `fileEncoding = "UTF-16LE"`, reading the first line, there are no 
errors. The BOM was removed automatically.

```{r one_row}
read.table(txt_file, nrows = 1, fileEncoding = "UTF-16LE", skipNul = TRUE) 
```

## Read more rows

We get an error when we try to read more rows.

```{r delim}
# Let's try reading more rows
try(read.table(txt_file, nrows = 5, fileEncoding = "UTF-16LE", skipNul = TRUE))
```

This error means that some rows are missing some values. We will look at whole 
lines to see what's going on. Let's start with the first five lines.

```{r read_whole_lines}
# Let's read in whole lines to see what they really look like
readLines(txt_file, encoding = "UTF-16LE", n = 5, skipNul = TRUE)
```

By inspection, we see the delimiter is tab (`\t`), there are blank lines, and 
no headers, with no specific value for "missingness", just the delimiter with
no value of any kind. When read into R, this would be the empty string ("").

## Set NA strings

Let's use what we know now about the file structure and try again using 
`na.strings = ""` to tell R that an empty string should be converted to `NA`.

```{r read_data_table}
df <- try(read.table(txt_file, header = FALSE, na.strings = "", 
                     fileEncoding = "UTF-16LE", skipNul = TRUE))
```

We will get the same error as before. That's because `read.table()` needs to know 
the delimiter with `sep = "\t"`.

## Setting the delimiter

We don't have to set `blank.lines.skip = TRUE`, since that's the default, but we 
do need to set `sep = "\t"`.

```{r read_data_table_sep}
# Read file as tab-delimited UTF-16LE, skipping nulls, with "" set to NA
df <- read.table(txt_file, header = FALSE, na.strings = "", sep = "\t", 
                 fileEncoding = "UTF-16LE", skipNul = TRUE)
```

If we use `read.delim()` instead of `read.table()`, we don't have to specify either 
`sep = "\t"` or `blank.lines.skip = TRUE` since the defaults will cover those.

```{r read_data_delim}
# Read file as whitepace-delimited UTF-16LE, skipping nulls, with "" set to NA
df <- read.delim(txt_file, header = FALSE, na.strings = "", 
                 fileEncoding = "UTF-16LE", skipNul = TRUE)
```

## Using `readr::read_delim`

We have been using `base`-R functions to read the file. We could use `readr` 
functions instead, but they do not handle the embedded nulls. So, you could 
work around that by processing with `base`-R's `readLines`  and `writeLines` 
first.

```{r, read_delim_readr, eval = FALSE}
# Do the same with readr::read_delim(), but after using readLines/writeLines
# since read_delim can't handle the embedded nulls
temp <- tempfile()
readLines(txt_file, encoding = "UTF-16LE", skipNul = TRUE) %>% writeLines(temp)
df <- read_delim(temp, delim = "\t", col_names = FALSE, na = "", 
                 name_repair = make.names)
res <- file.remove(temp)
```

We might also consider using `fread` from the `data.table` package, but it 
doesn't support UTF-16 encoding. So, again, we would be forced to read the 
file with something else first. It's hard to justify the extra trouble.

## Check dimensions

```{r check_dim}
# Check dimensions
dim(df)
```

That's a lot of columns! Maybe more than we need...

# Remove columns with no data

When we looked at the lines, we saw lots of empty values, so let's try to 
remove columns that contain no data, as they are just taking up space.

```{r initial_cleanup}
# Remove columns containing only NA values
df <- df %>% select(where(~ sum(is.na(.x)) < length(.x)))

# Check dimensions
dim(df)

# View data
df[1:5, 1:20]
```

## Remove rows with no drug data

By inspection, we see the drug data (name, quantity, etc.) starts at column V42. 
Some rows do not have any data from that column onwards, so we will remove those 
rows, as we don't need them.

```{r filter_rows}
# Find the numerical index of column V42 to use for subsetting
drug_start <- grep("V42", names(df))
drug_start

# Remove rows containing no drug data, e.g., where columns V42 onward are all NA
df <- df %>% filter(!if_all(all_of(drug_start:ncol(df)), ~ is.na(.)))

# Check dimensions
dim(df)
```

## Examine structure of drug data

When we look at the columns of drug data, we see a strange structure.

```{r examine_data_structure}
# View data
df %>% select(V42:V50) %>% head()
```

By inspection, we see columns containing drug data are in sets of 3:
drug_name drug_quantity drug_qualifier, example: "AMIKAC" " <= 16" "NOINTP".

## Reshape drug columns

We would prefer a structure where there were two colums per drug: one for the 
quantity and the other for the qualifier, which we will call "RIS". We want 
the two colums to be named after the drug, which is contained in the first 
of the three columns for each set. So, we will extract these first.

```{r rename_columns}
# Identify columns containing only a single abbreviated drug name (or NA)
drug_cols <- names(df)[drug_start:ncol(df)][drug_start:ncol(df) %% 3 == 0]
col_vals <- map(drug_cols, ~ table(df[, .x])) %>% setNames(drug_cols)

# Confirm all of these variables only contain 1 value matching name pattern
table(unlist(map(col_vals, length)))
table(unlist(map(col_vals, ~ grepl('^[A-Z]+$', names(.x)[1]))))

# Extract the drug names and column number
drugs <- sapply(col_vals, names)
drug_nums <- as.numeric(gsub('^V', '', drug_cols))

# Rename drug columns in sets of 3 columns for each drug: name, value, and RIS
drugs_df <- data.frame(drug = drugs, drug_num = drug_nums) %>%
  mutate(col1_old = paste0('V', drug_nums),
         col2_old = paste0('V', drug_nums + 1),
         col3_old = paste0('V', drug_nums + 2)) %>%
  mutate(col1_new = paste0(drug, '_NAME'), 
         col2_new = paste0(drug, '_VAL'),
         col3_new = paste0(drug, '_RIS'))
lookup <- with(drugs_df, 
               c(col1_old, col2_old, col3_old) %>%
      setNames(c(col1_new, col2_new, col3_new)))
df <- rename(df, any_of(lookup))
```

## Final cleanup

Now, all we need to do is remove extra spaces from the value columns, abbreviate 
the names of the drug value columns, and remove the original drug name columns. 
We won't worry about proving column names for the non-drug columns right now.

```{r final_cleanup}
# Clean up data
df <- df %>% 
  mutate(across(ends_with('_VAL'), ~ str_replace_all(.x, '\\s+', ' '))) %>%
  rename_with(~ str_remove(., '_VAL')) %>%
  select(-ends_with('_NAME'))

# View results
df[1:5, 15:20]
```

After saving the results to a CSV file, we are done.

```{r save_results}
# Save result as CSV
write_csv(df, csv_file)
```

## Appendix I: What's an "embedded null"?

A "embedded null" is a non-printing character (NUL) that is sometimes used to 
terminate a character string. You can see them using `readBin()`. This will 
show the hexadecimal equivalents of the characters. The nulls are the 00 values.

```{r see_nulls}
x <- readBin(txt_file, what = "raw", n = 16)
x
```

And we can "decode" the hexadecimal characters with `rawToChar()`, if we remove
the nulls first:

```{r decode}
paste(rawToChar(x[x!=0], multiple = TRUE), collapse = " ")
```

## Appendix II: What's UTF-16LE?

UTF-16 is a file encoding for 16-bit (2-byte) Unicode characters. We suspected we 
had an UTF-16LE file because of the two special bytes starting the file. These 
particular bytes, `FF FE` specify that the encoding is "little endian" (LE), giving us 
UTF-16LE encoding. 

- See: https://en.wikipedia.org/wiki/Endianness
- And: https://www.tutorialspoint.com/big-endian-and-little-endian

If we had a UTF-32LE file, it would have started with `FF FE 00 00`. If the file 
was UTF-8, it would either not have a BOM or the BOM would be `EF BB BF`. 

Because UTF-8 is represented with single bytes, there is no big or little "end". 
Endianness is to specify byte order, so you need more than one byte per character to 
have a byte order. So, the difference between "LE" and "BE" are just the order 
of the two bytes in each multibyte character. We can see the difference with 
an example that compares the bytes of LE and BE representing the character "1":

```{r le_vs_be}
x8 <- "1"
Encoding(x8) <- "UTF-8"
x16LE <- iconv(x8, "UTF-8", "UTF-16LE", toRaw = TRUE)
x16BE <- iconv(x8, "UTF-8", "UTF-16BE", toRaw = TRUE)
kable(
  t(c(x = x8, 'UTF-8' = charToRaw("1"), 'UTF-16LE' = x16LE, 'UTF-16BE' = x16BE))
  )
```

Those `00` values look like the dreaded "embedded nulls", right? But now we see 
those are valid after all. They are simply one of the two bytes of some UTF-16 
characters. That's why the warning said, *appears* to contain embedded nulls. 

## Appendix III: Good nulls and bad nulls

It turns out we actually need to "skip" the embedded nulls (`0x00 0x00`), or we 
will just read a few values.

```{r one_row_with_warnings_suppressed}
read.table(txt_file, nrows = 1, fileEncoding = "UTF-16LE") %>% suppressWarnings()
```

The reason is that we actually do have real embedded nulls after the third "cell".

```{r embedded_nulls}
readBin(txt_file, what = "raw", n = 100)
```

You can see that between some tabs (`\t` or `09 00`) we have `00 00` values. These
are the real embedded nulls because there are two of them together, so 16 bits of 
zeros.

I'm guessing these are the actual `NA` values of this file format. Since we don't 
have column headings, or any code book at all, we can't be sure, but if the tab 
is actually the delimiter, then this is how the empty "cells" are presented in 
this file. So, maybe we can try setting the null as a NA value:

```{r null_as_na, eval=FALSE}
try(read.table(txt_file, nrows = 1, fileEncoding = "UTF-16LE", na.strings = "\x00"))
```

```
Error: nul character not allowed (line 1)
```

Unfortunately we are not allowed to enter nulls here. So, we just skip them. 
That will leave the empty cells truly empty ("") and so we use the empty 
string as the NA string with `na.strings = ""`. 

At least now we have a better idea of what embedded nulls are, why 
they're there (probably), and why it's (probably) okay to skip them, at least in 
this situation.
